---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE, echo = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(dplyr)
library(aviandietdb)
library(scales)

dbsum = dbSummary()
```
# aviandietdb

<!-- badges: start -->
<!-- badges: end -->

**aviandietdb** provides access to the Avian Diet Database, which contains quantitative diet data for bird species as well as contextual information about where and when those data were collected.

The archived database currently includes **`r comma_format()(dbsum$numRecords)`** diet records for **`r dbsum$numSpecies`** bird species. This is a growing database, and while it is currently North American biased, we hope to continue adding avian diet data from around the world.

You can find the development version of the database [here](https://github.com/hurlbertlab/dietdatabase).

This R package provides a few simple functions for summarizing and querying the database.

## Installation

You can install **aviandietdb** from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ahhurlbert/aviandietdb")
```
## Database

Load the library and then the R data file that contains the raw data.

```{r}
library(aviandietdb)

data(dietdb)
```

Typing `?dietdb` will provide a description of the 42 data fields in this database.

A summary of the total number of records by species, and the total number of species with data by family is provided using `dbSummary()`.

```{r}
dbSummary()
```

## Summary functions

For now, we provide just three simple ways to summarize data from the Diet Database.

### speciesSummary()
The `speciesSummary()` function provides a summary of what kinds of data are available for the specified bird species, as well as a quantitative summary at the taxonomic level specified. For example,

```{r}
speciesSummary("Sharp-shinned hawk", by = "Class")
```

